---
title: "dioscRi quickstart"
author: "Elijah Willie"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{dioscRi quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

This vignette runs the dioscRi modelling stage end-to-end on the toy
dataset bundled with the package (10 synthetic samples, 5 markers,
5 cell types). The full BioHEART-CT walkthrough, including the
MMD-VAE normalisation step, is in `vignette("dioscRi_introduction")`.

```{r, message = FALSE}
library(dioscRi)
```

# Bundled inputs

The bundled `sample_*` objects already contain the pre-computed features
that the dioscRi modelling stage consumes. The MMD-VAE normalisation and
feature-construction steps that produced them are shown in the full
vignette; here we skip straight to modelling.

```{r}
data(sample_data_logit)     # logit-transformed cell-type proportions
data(sample_markerMeans)    # per-cell-type marker means
data(sample_groups)         # overlapping group structure
data(sample_clinicaldata)   # outcome + demographics

dim(sample_data_logit)
dim(sample_markerMeans)
length(sample_groups)
head(sample_clinicaldata)
```

# Build the design matrix

Concatenate proportions and per-cell-type marker means, scale, and
extract the binary outcome.

```{r}
X <- cbind(sample_data_logit, sample_markerMeans)
X <- scale(X)
y <- as.numeric(sample_clinicaldata$Outcome) - 1
```

# Fit the overlapping group lasso

`fitModel()` searches \(\alpha \in \{0.1, 0.2, \ldots, 1.0\}\) over a
fixed 10,000-point \(\lambda\) grid in \([0.05, 1]\), selecting the
\((\alpha, \lambda)\) that minimises BIC.

```{r}
fit <- fitModel(
  xTrain  = X,
  yTrain  = y,
  groups  = sample_groups,
  penalty = "grLasso",
  seed    = 1994
)
```

# Inspect the non-zero coefficients

```{r}
coefs <- coef(fit$fit)
coefs[coefs != 0]
```

# Predict and assess

`plotAUC()` returns the per-sample predictions and a ROC plot.
On a 10-sample toy dataset this is just to show the API; for real
held-out evaluation see the full vignette.

```{r}
auc_res <- plotAUC(fit = fit$fit, xTest = X, yTest = y, title = "Toy AUC =")
auc_res$auc
head(auc_res$preds)
auc_res$plot
```

# Where to go next

* `vignette("dioscRi_introduction")` — full BioHEART-CT walkthrough
  including reference-sample selection (`computeReferenceSample`),
  MMD-VAE training (`trainVAEModel`), unsupervised clustering
  (`trainCellTypeClassifier`), and downstream interpretation
  (`getSigFeatures`, `plotSigFeatures`, `visualiseModelTree`).
* The companion repository
  [`ecool50/dioscRi_manuscript`](https://github.com/ecool50/dioscRi_manuscript)
  contains the four datasets, scripts, and the v1.0.0 release used
  in the paper.

# Session info

```{r}
sessionInfo()
```
