---
title: "dioscRi with manually gated cell types"
author: "Elijah Willie"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
  %\VignetteIndexEntry{dioscRi with manually gated cell types}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r knitr-options, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE,
               fig.align = "center", fig.width = 7, fig.height = 5)
```

```{r load-cache, echo = FALSE}
.cache_file <- file.path(tools::R_user_dir("dioscRi", "cache"),
                         "bioheart_vignette_cache.rds")
.have_cache <- file.exists(.cache_file)
if (.have_cache) {
  .cache <- readRDS(.cache_file)
  sce_train          <- .cache$sce_train
  sce_test           <- .cache$sce_test
  clinicaldata_train <- .cache$clinicaldata_train
  clinicaldata_test  <- .cache$clinicaldata_test
  useMarkers         <- .cache$useMarkers
  nclust             <- .cache$nclust
}
```

# When to use this workflow

The [introduction vignette](dioscRi_introduction.html) defines cell types
from the data via FuseSOM clustering on the MMD-VAE latent space. If you
already have manually gated populations from domain experts (as in the
BioHEART-CT cohort), you can swap those in with three small changes to
the introduction pipeline:

1. Replace the unsupervised cluster label with the manual-gating column
   when training the LDA cell-type classifier and computing features.
2. Replace the data-driven tree (`generateTree()`) with a hand-built
   `data.tree::Node` reflecting the gating hierarchy.
3. Filter out cells whose manual label is `"other"` or `""` before
   computing features.

Everything else (MMD-VAE training, feature scaling, `fitModel()`,
`plotAUC()`, interpretation) is identical.

```{r show-cache-status, echo = FALSE, results = 'asis'}
if (.have_cache) {
  cat("This page reuses the MMD-VAE-trained SCE objects produced by the",
      "introduction vignette, so the chunks below run end-to-end and",
      "produce real plots.\n")
} else {
  cat("> The chunks below have `eval = FALSE` because the cache produced",
      "by the introduction vignette is not present. Render",
      "`vignette(\"dioscRi_introduction\")` first to populate the cache,",
      "then rebuild this article to see real plots.\n")
}
```

```{r setup-libs, eval = .have_cache, message = FALSE}
library(dioscRi)
library(SingleCellExperiment)
library(dplyr)
library(tidyr)
library(caret)
library(data.tree)
library(data.table)
library(ape)
library(ggtree)
```

# Step 1: transfer manual-gating labels to the test set

```{r transfer-mg, eval = .have_cache}
train_x <- reducedDim(sce_train, type = "VAE") |>
  mutate(cellTypes = as.factor(sce_train$mg_cell_type_distinct))
test_x  <- reducedDim(sce_test, type = "VAE")

sce_test$clusters_mg <- trainCellTypeClassifier(
  trainX = train_x, testX = test_x, model = "lda"
)
```

# Step 2: drop "other" / "" cells and compute features on the gating labels

```{r mg-features, eval = .have_cache}
sce_train <- sce_train[, !(sce_train$mg_cell_type_distinct %in% c("other", ""))]
sce_test  <- sce_test [, !(sce_test$clusters_mg          %in% c("other", ""))]
sce_test$clusters_mg <- droplevels(sce_test$clusters_mg)

# proportions
prop_logit_train <- computeFeatures(sce_train, featureType = "prop",
                                    cellTypeCol = "mg_cell_type_distinct",
                                    sampleCol = "sample_id",
                                    logit = TRUE, useMarkers = useMarkers)
prop_logit_test  <- computeFeatures(sce_test,  featureType = "prop",
                                    cellTypeCol = "clusters_mg",
                                    sampleCol = "sample_id",
                                    logit = TRUE, useMarkers = useMarkers)

# marker means
markerMeans_train <- computeFeatures(sce_train, featureType = "mean", assay = "norm",
                                     cellTypeCol = "mg_cell_type_distinct",
                                     sampleCol = "sample_id",
                                     logit = TRUE, useMarkers = useMarkers)
markerMeans_test  <- computeFeatures(sce_test,  featureType = "mean", assay = "norm",
                                     cellTypeCol = "clusters_mg",
                                     sampleCol = "sample_id",
                                     logit = TRUE, useMarkers = useMarkers)
```

# Step 3: build the gating hierarchy by hand

The manual hierarchy reflects the gating strategy used in the BioHEART-CT
study. `data.tree::Node` builds it; `findChildren(ggtree(...))` converts
it into the form `generateGroups()` expects.

```{r mg-tree, eval = .have_cache}
root <- Node$new("Root")
  myeloid <- root$AddChild("Myeloid_logit")
    myeloid$AddChild("14+ monos_logit")
    myeloid$AddChild("16+ monos_logit")
    myeloid$AddChild("pDCs_logit")
    myeloid$AddChild("CD141+ DCs_logit")
    myeloid$AddChild("CD1c+ DCs_logit")
  root$AddChild("NK_logit")
  root$AddChild("B cells_logit")
  CD3 <- root$AddChild("CD3_logit")
    CD3$AddChild("CD8hi_logit")
    CD3$AddChild("CD8lo_logit")
    CD4 <- CD3$AddChild("CD4_logit")
      CD4$AddChild("CD4+ Tconv_logit")
      CD4$AddChild("CD4+ Treg_logit")

nclust_mg <- ncol(prop_logit_train)
order_mg  <- as.phylo(root)$tip.label
tree_mg   <- findChildren(ggtree(as.phylo(root),
                                 ladderize = FALSE, layout = "dendrogram"))

groups <- generateGroups(tree = tree_mg, nClust = nclust_mg,
                         proportions = prop_logit_train,
                         means = markerMeans_train)
groups <- lapply(groups, sort)
```

# Step 4: fit and evaluate

Design-matrix construction, `fitModel()`, `plotAUC()`, and
`visualiseModelTree()` are unchanged from the introduction; only
`type = "MG"` is passed to the visualisation helpers so they use
the manual tree.

```{r mg-fit, eval = .have_cache}
y_train <- {
  rn <- rownames(prop_logit_train)
  y  <- factor(clinicaldata_train[match(rn, clinicaldata_train$sample_id), "Gensini_bin"])
  as.numeric(levels(y))[y]
}
y_test <- {
  rn <- rownames(prop_logit_test)
  y  <- factor(clinicaldata_test[match(rn, clinicaldata_test$sample_id), "Gensini_bin"])
  as.numeric(levels(y))[y]
}

X_train <- cbind(prop_logit_train, markerMeans_train) |> as.matrix()
scaleVals <- preProcess(X_train, method = "scale")
X_train   <- predict(scaleVals, X_train) |> as.matrix()
X_test    <- cbind(prop_logit_test, markerMeans_test) |> as.matrix()
X_test    <- predict(scaleVals, X_test)

fit <- fitModel(xTrain = X_train, yTrain = y_train,
                groups = groups, penalty = "grLasso", seed = 1994)
```

As in the unsupervised pipeline, the MMD-VAE's random initialisation
makes the test AUC a stochastic draw. Expect the combined-model test
AUC on the BioHEART-CT validation cohort in roughly [0.70, 0.85] across
re-runs; the published point estimate is 0.79.

## Training AUC

```{r mg-train-auc, eval = .have_cache}
plotAUC(fit$fit, X_train, y_train, title = "Train - AUC =")$plot
```

## Testing AUC

```{r mg-test-auc, eval = .have_cache}
plotAUC(fit$fit, X_test, y_test, title = "Test - AUC =")$plot
```

## Visualise the manual hierarchy

```{r mg-tree-plot, eval = .have_cache, fig.height = 8, fig.width = 12}
visualiseModelTree(fit$fit, tree = tree_mg, type = "MG",
                   trainingData = X_train,
                   nodesToRemove = c("Root", "Myeloid", "CD3", "CD4"),
                   title = "Proportions + Marker Means (manual gating)")
```

# Variants: proportions-only and means-only

The same overlap-lasso machinery can be applied to either feature family
on its own; the two variants below differ from the combined fit only in
the design matrix and the group structure.

## Proportions only

The groups come directly from the manual gating hierarchy
(`type = "prop"`).

```{r mg-fit-prop, eval = .have_cache}
X_train_p <- as.matrix(prop_logit_train)
X_test_p  <- as.matrix(prop_logit_test)

groups_p <- generateGroups(tree = tree_mg, nClust = nclust_mg,
                           proportions = prop_logit_train,
                           means = markerMeans_train, type = "prop")
groups_p <- lapply(groups_p, sort)

fit_p <- fitModel(xTrain = X_train_p, yTrain = y_train,
                  groups = groups_p, penalty = "grLasso", seed = 1994)
plotAUC(fit_p$fit, X_test_p, y_test, title = "Test - Proportions AUC =")$plot
```

## Means only

Each per-cell-type marker mean is in its own (singleton) group.

```{r mg-fit-mean, eval = .have_cache}
X_train_m <- as.matrix(markerMeans_train)
X_test_m  <- as.matrix(markerMeans_test)

groups_m <- as.list(seq_len(ncol(markerMeans_train)))

fit_m <- fitModel(xTrain = X_train_m, yTrain = y_train,
                  groups = groups_m, penalty = "grLasso", seed = 1994)
plotAUC(fit_m$fit, X_test_m, y_test, title = "Test - Means AUC =")$plot
```

# Session info

```{r}
sessionInfo()
```
