---
title: "Adding clinical covariates to the dioscRi model"
author: "Elijah Willie"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
  %\VignetteIndexEntry{Adding clinical covariates to the dioscRi model}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r knitr-options, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE,
               fig.align = "center", fig.width = 7, fig.height = 5)
```

```{r load-cache, echo = FALSE}
.cache_file <- file.path(tools::R_user_dir("dioscRi", "cache"),
                         "bioheart_vignette_cache.rds")
.have_cache <- file.exists(.cache_file)
if (.have_cache) {
  .cache <- readRDS(.cache_file)
  sce_train          <- .cache$sce_train
  sce_test           <- .cache$sce_test
  clinicaldata_train <- .cache$clinicaldata_train
  clinicaldata_test  <- .cache$clinicaldata_test
  useMarkers         <- .cache$useMarkers
  nclust             <- .cache$nclust
}
```

# When to use this workflow

The overlapping group lasso in dioscRi accommodates additional non-cytometry
covariates (age, sex, lab values, etc.) by appending them to the scaled
design matrix and adding a corresponding group to the group structure. On
the BioHEART-CT validation cohort, the published model that combines
cytometry features with age and sex reaches **AUC = 0.87**, up from 0.83
with cytometry alone.

This vignette starts from the cached SCE objects produced by the
[introduction vignette](dioscRi_introduction.html); render that one
first to populate the cache.

```{r show-cache-status, echo = FALSE, results = 'asis'}
if (.have_cache) {
  cat("The chunks below reuse the cache and run end-to-end.\n")
} else {
  cat("> The cache produced by the introduction vignette is not present;",
      "render `vignette(\"dioscRi_introduction\")` first to populate it.",
      "Chunks below have `eval = FALSE` until the cache is available.\n")
}
```

```{r setup-libs, eval = .have_cache, message = FALSE}
library(dioscRi)
library(SingleCellExperiment)
library(dplyr)
library(tidyr)
library(caret)
```

# Re-derive features from the cached SCEs

The intro vignette saves the post-VAE SCEs along with their cluster
labels. We can recompute the proportion + marker-mean features quickly.

```{r features, eval = .have_cache}
prop_logit_train <- computeFeatures(sce_train, featureType = "prop",
                                    cellTypeCol = "clusters_norm",
                                    sampleCol = "sample_id",
                                    logit = TRUE, useMarkers = useMarkers)
prop_logit_test  <- computeFeatures(sce_test,  featureType = "prop",
                                    cellTypeCol = "clusters_norm",
                                    sampleCol = "sample_id",
                                    logit = TRUE, useMarkers = useMarkers)

markerMeans_train <- computeFeatures(sce_train, featureType = "mean",
                                     cellTypeCol = "clusters_norm",
                                     sampleCol = "sample_id",
                                     logit = TRUE, useMarkers = useMarkers)
markerMeans_test  <- computeFeatures(sce_test,  featureType = "mean",
                                     cellTypeCol = "clusters_norm",
                                     sampleCol = "sample_id",
                                     logit = TRUE, useMarkers = useMarkers)
```

# One-hot encode sex and align rows

`Gender` is binary so we expand it to a column `GenderM` (1 = male,
0 = female). The age column is already numeric.

```{r encode-clinical, eval = .have_cache}
clinicaldata_train <- clinicaldata_train |>
  mutate(GenderM = as.integer(Gender == "M"))
clinicaldata_test  <- clinicaldata_test  |>
  mutate(GenderM = as.integer(Gender == "M"))

# Match clinical rows to feature rows (rownames are sample ids)
match_train <- match(rownames(prop_logit_train), clinicaldata_train$sample_id)
match_test  <- match(rownames(prop_logit_test),  clinicaldata_test$sample_id)

clinical_train <- clinicaldata_train[match_train, c("Age", "GenderM")] |>
  as.matrix()
clinical_test  <- clinicaldata_test [match_test,  c("Age", "GenderM")] |>
  as.matrix()
```

# Build the design matrix with clinical columns

The cytometry features are z-scaled; the clinical columns are appended
*after* scaling so that their unscaled values pass through to the lasso.

```{r design-matrix, eval = .have_cache}
y_train <- {
  y <- factor(clinicaldata_train[match_train, "Gensini_bin"])
  as.numeric(levels(y))[y]
}
y_test <- {
  y <- factor(clinicaldata_test[match_test, "Gensini_bin"])
  as.numeric(levels(y))[y]
}

# scale cytometry features only
cytof_train <- cbind(prop_logit_train, markerMeans_train) |> as.matrix()
scaleVals <- preProcess(cytof_train, method = "scale")
cytof_train <- predict(scaleVals, cytof_train) |> as.matrix()
cytof_test  <- cbind(prop_logit_test, markerMeans_test) |> as.matrix()
cytof_test  <- predict(scaleVals, cytof_test) |> as.matrix()

# append age + sex
X_train <- cbind(cytof_train, clinical_train)
X_test  <- cbind(cytof_test,  clinical_test)

dim(X_train); dim(X_test)
```

# Extend the group structure

The cytometry groups come from the unsupervised hierarchy; we add one
additional singleton group covering the two clinical columns so they
enter the model as their own block.

```{r groups, eval = .have_cache}
trees  <- generateTree(features = prop_logit_train, method = "ward")
tree   <- trees$tree
groups <- generateGroups(tree = tree, nClust = nclust,
                         proportions = prop_logit_train,
                         means = markerMeans_train)

clin_idx <- (ncol(cytof_train) + 1):ncol(X_train)  # Age, GenderM positions
groups[[length(groups) + 1]] <- clin_idx
groups <- lapply(groups, sort)
```

# Fit and evaluate

```{r fit, eval = .have_cache}
fit <- fitModel(xTrain = X_train, yTrain = y_train,
                groups = groups, penalty = "grLasso", seed = 1994)
```

## Training AUC

```{r train-auc, eval = .have_cache}
plotAUC(fit$fit, X_train, y_train, title = "Train - AUC =")$plot
```

## Testing AUC

```{r test-auc, eval = .have_cache}
plotAUC(fit$fit, X_test, y_test, title = "Test (cytometry + Age + Sex) - AUC =")$plot
```

The manuscript reports AUC = 0.87 for this combination on the validation
cohort. Expect single-run results within roughly [0.78, 0.90] because of
MMD-VAE / FuseSOM stochasticity.

# Inspect the non-zero clinical coefficients

```{r coefs, eval = .have_cache}
coefs <- coef(fit$fit)
coefs[c("Age", "GenderM")]
```

# Session info

```{r}
sessionInfo()
```
