---
title: "dioscRi package manual"
author: "Elijah Willie"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
  %\VignetteIndexEntry{dioscRi package manual}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r knitr-options, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
# opts_chunk$set(fig.align = 'center', fig.width = 6, fig.height = 5, dev = 'png')
```

# Installation
```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("https://github.com/ecool50/dioscRi/")
library(dioscRi)
```

# Introduction

`dioscRi` is an R package designed to provide an end-to-end solution for identifying significant features in single-cell or multiplexed imaging data. Leveraging hierarchical clustering, overlapping group lasso, and downstream interpretative tools, dioscRi allows researchers to identify and understand the biological relevance of features, offering a robust framework for hypothesis generation and data-driven insights.

# Overview of the dioscRi workflow

The dioscRi pipeline comprises several distinct yet interconnected steps, ensuring a smooth analytical workflow for feature identification and downstream interpretation:

<img src=https://raw.githubusercontent.com/ecool50/dioscRi/main/inst/dioscRi_overview.png align="middle" height="400" width="1200">

## Summary of the **dioscRi** Pipeline

- **Transferable Normalization**:
  - A neural network is trained on the training samples to remove technical variability while preserving biological heterogeneity.
  - This step ensures consistent data normalization for downstream analysis.

- **Cell Typing**:
  - Cell populations are identified using clustering, manual gating, or reference-based methods.
  - This step establishes distinct cell type labels for each sample.

- **Cell Type Grouping**:
  - Hierarchical relationships between cell types are defined using tools like **treekoR** or through manual annotations.
  - Groups are structured for interpretability and predictive modeling.

- **Prediction and Interpretation**:
  - Hierarchical group lasso models are used to make predictions based on grouped cell types or features.
  - The model provides interpretable outputs to understand key contributing features or cell populations.

- **New Sample Normalization**:
  - The trained normalization model is applied to normalize new samples, ensuring compatibility with the training data.

- **Cell Type Classification**:
  - Cell types in new samples are classified based on the pre-defined cell type structure and trained model.

- **New Sample Classification**:
  - Patient-level predictions are generated for new samples, providing probabilities or classifications (e.g., healthy vs. diseased).
  - This final step enables the application of the model for clinical or experimental outcomes.

# Load in required libraries
```{r}
rm(list = ls())
gc()
suppressPackageStartupMessages({
    library(dioscRi)
library(keras3)
library(tidyverse)
library(tensorflow)
library(data.table)
library(caret)
library(SingleCellExperiment)
library(FuseSOM)
})

set_random_seed(1994)
```

# Set the markers
```{r}
useMarkers <- c('HLA_DR', 'CD3', 'CD4', 'CD8a', 'CD25', 'CD127', 'FoxP3', 'CD27',
                'KLRG1', 'CD56', 'CD45RO', 'CD45RA', 'CD192_CCR2', 'CD194_CCR4',
                'CD196_CCR6',
                'CD39', 'CD38', 'Ki67', 'CD183_CXCR3', 'CCR7', 'CD19', 'CD20',
                'IgD', 'CD14', 'CD304', 'CD141', 'CD1c_PE')

allMarkers <- c('HLA_DR', 'CD3', 'CD4', 'CD8a', 'CD25', 'CD127', 'FoxP3', 'CD27',
                'KLRG1', 'CD56', 'CD45RO', 'CD45RA', 'CD192_CCR2', 'CD194_CCR4',
                'CD196_CCR6',
                'CD39', 'CD38', 'Ki67', 'CD183_CXCR3', 'CCR7', 'CD19', 'CD20',
                'IgD', 'CD14', 'CD304', 'CD141', 'CD1c_PE', "CD11b", "CD253_TRAIL",
                "CD34", "CD61", "CD11c", "eNOS", "LOX_1", "CD86", "CD16", "CD45_106",
                "CD45_108", "P2X7", "NOX5")
```


# Read in the datasets
## Discovery study for training
```{r}
# study 4
set.seed(1994)
df_train <- fread('../../raw_data/bioheart_ct_cytof_data_b4_mg.csv',
              nThread = 7) %>%
  as.data.frame()
```


# Transform the training dataset using the arsinh function with cofactor of 5
```{r}
df_train[, useMarkers]  <- cyCombine::transform_asinh(df_train[, useMarkers], 
                                                markers = useMarkers, derand = F)

colnames(df_train)[colnames(df_train) %in% useMarkers] <- 
  gsub("_", "-", colnames(df_train)[colnames(df_train) %in% useMarkers])

```

## Validation study for testing
```{r}
set.seed(1994)
df_test <- fread('../../Study_3_2019/all_batches_processed_mg_10K.csv',
              nThread = 7) %>% 
  as.data.frame()

df_test[, useMarkers]  <- cyCombine::transform_asinh(df_test[, useMarkers],
                                                  markers = useMarkers, derand = F)

colnames(df_test)[colnames(df_test) %in% useMarkers] <- gsub("_", "-", colnames(df_test)[colnames(df_test) %in% useMarkers])
```

# Scale training and testing data for deep learning
```{r}
useMarkers <- gsub("_", "-", useMarkers)
allMarkers <- gsub("_", "-", allMarkers)

preProcValues <- preProcess(df_train[, useMarkers], method = c("range"))

df_train[, useMarkers] <- predict(preProcValues, df_train[, useMarkers])

df_test[, useMarkers] <- predict(preProcValues, df_test[, useMarkers])
```
 
# Get the reference sample for training
```{r}
res_ind <- computeReferenceSample(df_train, useMarkers, N = 2)
```

# Get the training and validation samples for the VAE model
```{r}
train_dat <- df_train[which(df_train$sample_id %in% c(res_ind$topNSamples)), ]
val_dat <- df_train[which(df_train$sample_id %in% c(res_ind$bottomNSamples)), ]
```


# Run the MMD-VAE on the training data
```{r}
batch_size = 32L
vae_model <- trainVAEModel(trainData = train_dat[, useMarkers], batchSize = batch_size, 
                             useMarkers = useMarkers, epochs = 100, lambda = 0.01,
                             valData = val_dat[, useMarkers])
```

# Use the trained VAE model to decode the training data
```{r}
train_decoded <- decodeSamples(newSamples = as.matrix(df_train[, useMarkers]), 
                                vae = vae_model$vae)
df_train_norm <- train_decoded[[1]]

colnames(df_train_norm) <- useMarkers
```

# Use the trained VAE model to decode the testing data
```{r, eval=T}
test_decoded <- decodeSamples(newSamples = as.matrix(df_test[, useMarkers]), 
                               vae = vae_model$vae)
df_test_norm <- test_decoded[[1]]

colnames(df_test_norm) <- useMarkers
```


# Create SCE object fort the training data
```{r}
sce_train_norm <- SingleCellExperiment(assays = list(norm = t(df_train_norm[, useMarkers])
),
colData = df_train %>% dplyr::select(-useMarkers))
reducedDims(sce_train_norm) <- list(VAE=train_decoded$encoded %>%
                                as.matrix() %>%
                                as.data.frame())
```

# Create SCE object for the testing data
```{r}
sce_test_norm <- SingleCellExperiment(assays = list(norm = t(df_test_norm[, useMarkers])
),
colData = df_test %>% dplyr::select(-useMarkers))
reducedDims(sce_test_norm) <- list(VAE=test_decoded$encoded %>%
                                as.matrix() %>%
                                as.data.frame())


```

## Use FuseSOM to cluster the normalised training data into 11 clusters
```{r, eval=T}
sce_train_norm$clusters_norm <- NULL
nclust <- 11
sce_train_norm <- runFuseSOM(sce_train_norm, numClusters = nclust, assay = 'norm', 
                       verbose = FALSE, clusterCol = 'clusters_norm')
```

# Parse Clinical data for the training data
```{r}
clinicaldata <- colData(sce_train_norm) %>%
  as.data.frame() %>%
  dplyr::select(sample_id, Gensini, Gender, Age) %>%
  distinct()
clinicaldata$Gensini_bin <- factor(if_else(clinicaldata$Gensini > 0, 1, 0))
```

# Parse Clinical data for testing data
```{r}

clinicaldata_test <- colData(sce_test_norm) %>%
  as.data.frame() %>%
  dplyr::select(sample_id, Gensini_bin) %>%
  distinct()
```

# Train cell type identification model using the training data
We will be using the the embeddings from the VAE model to train a linear discriminant model to predict the clusters for the testing dat
```{r}
train_x <- reducedDim(sce_train_norm, type = "VAE") %>%
  mutate(cellTypes = as.factor(sce_train_norm$clusters_norm))

test_x <- reducedDim(sce_test_norm, type = "VAE")
```

## Fit the LDA model
```{r}
test_clusters <- trainCellTypeClassifier(trainX = train_x, 
                                         testX = test_x, model = 'lda')

sce_test_norm$clusters_norm <- test_clusters
```

# Compute training features
## Logit of cell type proportions
```{r}
prop_logit <- computeFeatures(sce = sce_train_norm, featureType = 'prop', 
                              cellTypeCol = 'clusters_norm', sampleCol = 'sample_id', 
                              logit = T, useMarkers = useMarkers)
row_names <- rownames(prop_logit)
condition <- factor(clinicaldata[match(row_names,clinicaldata$sample_id),
                                      "Gensini_bin"])
```

## Means
```{r, eval=T}
markerMeanCellType <- computeFeatures(sce = sce_train_norm, featureType = 'mean', 
                              cellTypeCol = 'clusters_norm', sampleCol = 'sample_id', 
                              logit = T, useMarkers = useMarkers)
```


# Compute features on test data
## Logit of cell type proportions
```{r}
prop_test_logit <- computeFeatures(sce = sce_test_norm, featureType = 'prop', 
                              cellTypeCol = 'clusters_norm', sampleCol = 'sample_id', 
                              logit = T, useMarkers = useMarkers)

row_names_test <- rownames(prop_test_logit)
condition_test <- factor(clinicaldata_test[match(row_names_test,clinicaldata_test$sample_id),
                                      "Gensini_bin"])
```

## Means
```{r, eval=T}
markerMeanCellType_test <- computeFeatures(sce = sce_test_norm, featureType = 'mean', 
                              cellTypeCol = 'clusters_norm', sampleCol = 'sample_id', 
                              logit = T, useMarkers = useMarkers)
```


# Generate grouping structures using the clusters
```{r}
trees <- generateTree(features = prop_logit, method = "ward")
tree <- trees$tree
order <- trees$order
groups <- generateGroups(tree = tree, nClust = nclust, 
                         proportions = prop_logit, means = markerMeanCellType)
```

# Process training and testing data
## Training
```{r}
# Combine all input data matrices
X_train <- cbind(prop_logit, markerMeanCellType) %>% 
  as.matrix()
  
  # Scale the combined data
scaleVals <- preProcess(X_train, method = c('scale'))
X_train <- predict(scaleVals, X_train) %>%
    as.matrix()
  
  # Extract response variable
y_train <- as.numeric(levels(condition))[condition]
```

## Testing
```{r}
X_test <- cbind(prop_test_logit, markerMeanCellType_test) %>%
  as.matrix()

X_test <- predict(scaleVals, X_test) %>%
    as.matrix()

y_test <- as.numeric(levels(condition_test))[condition_test]
```

## Fit Overlap group lasso model on the training data
```{r}
fit <- fitModel(xTrain = X_train, yTrain = y_train, groups = groups, 
                penalty = 'grLasso', seed = 1994)
```

```{r}
train_auc <- plotAUC(fit = fit$fit, xTest = X_train, yTest = y_train, title = "Study 3 - All AUC =")
train_auc$plot
```

## Predict on test data
```{r}
test_auc <- plotAUC(fit = fit$fit, xTest = X_test, yTest = y_test, title = "Study 3 - All AUC =")
test_auc$plot

```

# Visualise Resulting Tree
```{r, fig.height=20, fig.width=40}
visualiseModelTree(fit = fit$fit, tree = tree, type = "cluster",
                   trainingData = X_train, title = "Unsupervised Clustering")
```