---
title: "dioscRi package manual"
author: "Elijah Willie"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
  %\VignetteIndexEntry{dioscRi package manual}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r knitr-options, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', fig.width = 6, fig.height = 5, dev = 'png')
```

# Installation
```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("https://github.com/ecool50/dioscRi/")
library(dioscRi)
```

# Introduction

`dioscRi` is an R package designed to provide an end-to-end solution for identifying significant features in single-cell or multiplexed imaging data. Leveraging hierarchical clustering, overlapping group lasso, and downstream interpretative tools, dioscRi allows researchers to identify and understand the biological relevance of features, offering a robust framework for hypothesis generation and data-driven insights.

# Overview of the dioscRi workflow

The dioscRi pipeline comprises several distinct yet interconnected steps, ensuring a smooth analytical workflow for feature identification and downstream interpretation:

 